Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-02-28 article-models-tt.tex
proof of currying complete

In \autoref{extension-on-terms} we come back to the operation of extension, but
now for terms. The rules in this section explain what it means to be a term of
an extended family, namely to be an extended term. The compatibility rules for
extended terms are then similar as the compatibility rules that have appeared
earlier: there is a compatibility rule for the extension of an extension, for
the weakening of an extended term, for the substutution of an extended term by
an arbitrary term and for substitution by an extended term. The last set of
compatibility rules is a form of the Curry-Howard isomorphism -- this famous
correspondence naturally comes out of our scheme of compatibility rules.

@@

\subsubsection{Weakening is compatible with itself}\label{comp-ww}
We state judgmental equality rules expressing
that weakening is compatible with itself. These rules state that the following
diagram commutes given any two families $A$ and $B$ in context $\Gamma$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\Gamma}{\blank} \ar{r}{C\mapsto\ctxwk{B}{C}} \ar{d}[swap]{C\mapsto\ctxwk{A}{C}} & \jfam{\ctxext{\Gamma}{B}}{\blank} \ar{d}{Q\mapsto\ctxwk{A}{Q}}\\
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{r}[swap]{P\mapsto\ctxwk{{A}{B}}{P}} & \jfam{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\blank}
\end{tikzcd}
\end{equation*}
Thus, we get the following set of inference rules:
\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jfam{\ctxext{\Gamma}{C}}{R}}
          {\jfameq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{A}{{B}{R}}}
            {\ctxwk{{A}{B}}{{A}{R}}}}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jterm{\ctxext{\Gamma}{C}}{R}{t}}
          {\jtermeq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{{A}{B}}{{A}{R}}}{\ctxwk{A}{{B}{t}}}{\ctxwk{{A}{B}}{{A}{t}}}}
            \label{comp-ww-inf4}
\end{align}

@@

In the following inference rules we assume that
$\jfam{\Gamma}{A}$ and $\jfam{\ctxext{\Gamma}{A}}{P}$. 
\begin{align}
& \inference{\jfam{\ctxext{\Gamma}{B}}{Q}}
    {\jfameq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{\ctxext{A}{P}}{Q}}{\ctxwk{P}{{A}{Q}}}}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}
    {\jtermeq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{P}{{A}{Q}}}{\ctxwk{\ctxext{A}{P}}{g}}{\ctxwk{P}{{A}{g}}}}
\end{align}

@@

\subsubsection{Substitution is compatible with weakening}\label{comp-sw}
The rules asserting the compatibility of substitution with weakening assert
that the following diagram commutes for any $\jterm{\Gamma}{A}{x}$ and any
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{d}[swap]{Q\mapsto\subst{x}{Q}} \ar{r}{Q\mapsto\ctxwk{P}{Q}} & \jfam{\ctxext{{\Gamma}{A}}{P}}{\blank} \ar{d}{R\mapsto\subst{x}{R}}\\ 
\jfam{\Gamma}{\blank} \ar{r}[swap]{B\mapsto\ctxwk{\subst{x}{P}}{B}} & \jfam{\ctxext{\Gamma}{\subst{x}{P}}}{\blank}
\end{tikzcd}
\end{equation*}
We plug in an extra layer of families to cover the most general case at once.
In the following inference rules we assume that $\jterm{\Gamma}{A}{x}$ and
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{align}
& \inference
  {\jfam{\ctxext{{\Gamma}{A}}{Q}}{R}}
  {\jfameq{\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}{\subst{x}{\ctxwk{P}{R}}}{\ctxwk{\subst{x}{P}}{\subst{x}{R}}}}\\
& \inference
    {\jterm{\ctxext{{\Gamma}{A}}{Q}}{R}{h}}
    {\jtermeq
      {\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}
      {\subst{x}{\ctxwk{P}{R}}}
      {\subst{x}{\ctxwk{P}{h}}}
      {\ctxwk{\subst{x}{P}}{\subst{x}{h}}}}
\end{align}

@@

\begin{align}
&\inference{\jfam{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
{\jfameq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{{x}{f}}{{x}{R}}}}\label{comp-ss-3}\\
&\inference{\jterm{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}{h}}
{\jtermeq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{x}{{f}{h}}}{\subst{{x}{f}}{{x}{h}}}}\label{comp-ss-4}
\end{align}

@@

\begin{align}
& \inference{\jfam{\ctxext{{\Gamma}{B}}{Q}}{R}}{\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{R}}}}\\
& \inference{\jterm{\ctxext{{\Gamma}{B}}{Q}}{R}{h}}{\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\ctxwk{A}{\subst{y}{h}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{h}}}}
\end{align}

@@

\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jterm{\Gamma}{A}{x}}{\jfameq{\Gamma}{\subst{x}{\ctxwk{A}{B}}}{B}}\label{defn-ws-1}\\
& \inference{\jterm{\Gamma}{A}{x}\qquad\jterm{\Gamma}{B}{y}}{\jtermeq{\Gamma}{B}{\subst{x}{\ctxwk{A}{y}}}{y}}\label{defn-ws-2}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\ctxext{\Gamma}{B}}{Q}\qquad\jterm{\Gamma}{A}{x}}{\jfameq{\ctxext{\Gamma}{B}}{\subst{x}{\ctxwk{A}{Q}}}{Q}}\label{defn-ws-3}\\
& \inference{\jterm{\Gamma}{A}{x}\qquad\jterm{\ctxext{\Gamma}{B}}{Q}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{Q}{\subst{x}{\ctxwk{A}{g}}}{g}}\label{defn-ws-4}
\end{align}

@@

\subsubsection{Identity functions}
Without a rule explicitly asserting the existence of an identity morphism we don't
get one, hence we do that here. The identity morphism is a term which introduced
in ordinary type theory via the variable rule. The variable rule is a bit more
general: it asserts that
\begin{equation*}
\jterm{\Gamma,\,x_1:A_1,\ldots,\,x_n:A_n}{A_i}{x_i}
\end{equation*}
for every $1\leq i\leq n$. Thus, it establishes the projections. In our setting,
we get the projections from the identity morphisms together with weakening. We
already have weakening, so here it suffices to introduce the identity morphisms.
\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}}
& & \inference{\jfameq{\Gamma}{A}{A'}}{\jtermeq{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}{\idfunc[A']}}
\end{align}
Identity functions are determined by their behavior with respect to substitution combined with
weakening. The identity functions will also be subject to compatibility rules.
\begin{align}
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{x}{\idfunc[A]}}{x}}\label{idfunc-subst-defn}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{A}}{\subst{\idfunc[A]}{\ctxwk{A}{P}}}{P}}\label{idfunc-wk-defn}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\idfunc[A]}{\ctxwk{A}{f}}}{f}}\label{idfunc-precomp} \\
& \inference{\jterm{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{\subst{g}{\ctxwk{B}{\idfunc[A]}}}{g}}\label{idfunc-postcomp}
\end{align}

@@

@@

\begin{proof}
To prove the first judgmental equality, note that
\begin{align*}
w & \jdeq \subst{w}{\idfunc[\ctxext{A}{P}]} \tag{by \autoref{idfunc-subst-defn}}\\
& \jdeq \subst{w}{\subst{\idfunc[P]}{{\ctxwk{P}{\idfunc[A]}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}
  \tag{by \autoref{idfunc-ext-comp}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{\subst{w}{{\ctxwk{P}{\idfunc[A]}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}\tag{by \autoref{comp-ss-4}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\subst{w}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}\tag{by \autoref{comp-ss-4}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\subst{w}{\ctxwk{\ctxext{A}{P}}{\idfunc[\ctxext{A}{P}]}}}}\tag{by \autoref{idfunc-wk-comp}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\idfunc[\ctxext{A}{P}]}}\tag{by \autoref{defn-ws-2}}\\
& \jdeq \ctxext{\cprojfst{A}{P}{w}}{\cprojsnd{A}{P}{w}}\tag{by definition}
\end{align*}
\end{proof}

@@

\begin{lem}
When we substitute an extended term $\ctxext{f}{g}$ of $\ctxext{P}{Q}$ by a term
$x$ of $A$, the term that we get is $\ctxext{\subst{x}{f}}{\subst{x}{g}}$.
More precisely, the following inference rules are valid:
\begin{align*}
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}\quad\jterm{\ctxext{\Gamma}{A}}{\subst{f}{Q}}{g}}{\jtermeq{\Gamma}{\ctxext{\subst{x}{P}}{\subst{x}{Q}}}{\subst{x}{\ctxext{f}{g}}}{\ctxext{\subst{x}{f}}{\subst{x}{g}}}}\\
& \inference{\jterm{\ctxext{{\Gamma}{A}}{P}}{Q}{g}\quad\jterm{\ctxext{{\Gamma}{A}}{P}}{\subst{g}{R}}{t}}{\jtermeq{\ctxext{\Gamma}{\subst{x}{P}}}{\ctxext{\subst{x}{Q}}{\subst{x}{R}}}{\subst{x}{\ctxext{g}{t}}}{\ctxext{\subst{x}{g}}{\subst{x}{t}}}}
\end{align*}
\end{lem}
Finally, when we substitute by an extended term we get an equal result as when we
substitute two consecutive times. This looks like the Curry-Howard isomorphism.

@@

>>>>

In \autoref{extension-on-terms} we come back to the operation of extension, but
now for terms. The rules in this section explain what it means to be a term of
an extended family, namely to be an extended term. The compatibility rules for
extended terms are then similar as the compatibility rules that have appeared
earlier: there is a compatibility rule for the extension of an extension, for
the weakening of an extended term, for the substutution of an extended term by
an arbitrary term and for substitution by an extended term. The last set of
compatibility rules is a form of currying-uncurrying -- this famous
correspondence naturally comes out of our scheme of compatibility rules.

@@

\subsubsection{Weakening is compatible with itself}\label{comp-ww}
We state judgmental equality rules expressing
that weakening is compatible with itself. These rules state that the following
diagram commutes given any two families $A$ and $B$ in context $\Gamma$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\Gamma}{\blank} \ar{r}{C\mapsto\ctxwk{B}{C}} \ar{d}[swap]{C\mapsto\ctxwk{A}{C}} & \jfam{\ctxext{\Gamma}{B}}{\blank} \ar{d}{Q\mapsto\ctxwk{A}{Q}}\\
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{r}[swap]{P\mapsto\ctxwk{{A}{B}}{P}} & \jfam{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\blank}
\end{tikzcd}
\end{equation*}
Thus, we get the following set of inference rules:
\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jfam{\ctxext{\Gamma}{C}}{R}}
          {\jfameq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{A}{{B}{R}}}
            {\ctxwk{{A}{B}}{{A}{R}}}}\label{comp-ww-f}\\
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\Gamma}{B}\qquad\jterm{\ctxext{\Gamma}{C}}{R}{t}}
          {\jtermeq{\ctxext{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{C}}}}
            {\ctxwk{{A}{B}}{{A}{R}}}{\ctxwk{A}{{B}{t}}}{\ctxwk{{A}{B}}{{A}{t}}}}
            \label{comp-ww-t}
\end{align}

@@

In the following inference rules we assume that
$\jfam{\Gamma}{A}$ and $\jfam{\ctxext{\Gamma}{A}}{P}$. 
\begin{align}
& \inference{\jfam{\ctxext{\Gamma}{B}}{Q}}
    {\jfameq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{\ctxext{A}{P}}{Q}}{\ctxwk{P}{{A}{Q}}}}\label{comp-ew-f}\\
& \inference{\jterm{\ctxext{\Gamma}{B}}{Q}{g}}
    {\jtermeq{\ctxext{{{\Gamma}{A}}{P}}{\ctxwk{P}{{A}{B}}}}{\ctxwk{P}{{A}{Q}}}{\ctxwk{\ctxext{A}{P}}{g}}{\ctxwk{P}{{A}{g}}}}\label{comp-ew-t}
\end{align}

@@

\subsubsection{Substitution is compatible with weakening}\label{comp-sw}
The rules asserting the compatibility of substitution with weakening assert
that the following diagram commutes for any $\jterm{\Gamma}{A}{x}$ and any
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\jfam{\ctxext{\Gamma}{A}}{\blank} \ar{d}[swap]{Q\mapsto\subst{x}{Q}} \ar{r}{Q\mapsto\ctxwk{P}{Q}} & \jfam{\ctxext{{\Gamma}{A}}{P}}{\blank} \ar{d}{R\mapsto\subst{x}{R}}\\ 
\jfam{\Gamma}{\blank} \ar{r}[swap]{B\mapsto\ctxwk{\subst{x}{P}}{B}} & \jfam{\ctxext{\Gamma}{\subst{x}{P}}}{\blank}
\end{tikzcd}
\end{equation*}
We plug in an extra layer of families to cover the most general case at once.
In the following inference rules we assume that $\jterm{\Gamma}{A}{x}$ and
$\jfam{\ctxext{\Gamma}{A}}{P}$:
\begin{align}
& \inference
  {\jfam{\ctxext{{\Gamma}{A}}{Q}}{R}}
  {\jfameq{\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}{\subst{x}{\ctxwk{P}{R}}}{\ctxwk{\subst{x}{P}}{\subst{x}{R}}}}\label{comp-sw-f}\\
& \inference
    {\jterm{\ctxext{{\Gamma}{A}}{Q}}{R}{h}}
    {\jtermeq
      {\ctxext{{\Gamma}{\subst{x}{P}}}{\subst{x}{\ctxwk{P}{Q}}}}
      {\subst{x}{\ctxwk{P}{R}}}
      {\subst{x}{\ctxwk{P}{h}}}
      {\ctxwk{\subst{x}{P}}{\subst{x}{h}}}}\label{comp-sw-t}
\end{align}

@@

\begin{align}
&\inference{\jfam{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}}
{\jfameq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{{x}{f}}{{x}{R}}}}\label{comp-ss-f}\\
&\inference{\jterm{\ctxext{{{\Gamma}{A}}{P}}{Q}}{R}{h}}
{\jtermeq{\ctxext{\Gamma}{\subst{x}{{f}{Q}}}}{\subst{x}{{f}{R}}}{\subst{x}{{f}{h}}}{\subst{{x}{f}}{{x}{h}}}}\label{comp-ss-t}
\end{align}

@@

\begin{align}
& \inference{\jfam{\ctxext{{\Gamma}{B}}{Q}}{R}}{\jfameq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{R}}}}\label{comp-ws-f}\\
& \inference{\jterm{\ctxext{{\Gamma}{B}}{Q}}{R}{h}}{\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{\subst{y}{Q}}}}{\ctxwk{A}{\subst{y}{R}}}{\ctxwk{A}{\subst{y}{h}}}{\subst{\ctxwk{A}{y}}{\ctxwk{A}{h}}}}\label{comp-ws-t}
\end{align}

@@

\begin{align}
& \inference{\jfam{\Gamma}{A}\qquad\jfam{\ctxext{\Gamma}{B}}{Q}\qquad\jterm{\Gamma}{A}{x}}{\jfameq{\ctxext{\Gamma}{B}}{\subst{x}{\ctxwk{A}{Q}}}{Q}}\label{defn-ws-3}\\
& \inference{\jterm{\Gamma}{A}{x}\qquad\jterm{\ctxext{\Gamma}{B}}{Q}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{Q}{\subst{x}{\ctxwk{A}{g}}}{g}}\label{defn-ws-4}
\end{align}

@@

\subsubsection{Identity functions}
Without a rule explicitly asserting the existence of an identity morphism we don't
get one, hence we do that here. The identity morphism is a term which introduced
in ordinary type theory via the variable rule. The variable rule is a bit more
general: it asserts that
\begin{equation*}
\jterm{\Gamma,\,x_1:A_1,\ldots,\,x_n:A_n}{A_i}{x_i}
\end{equation*}
for every $1\leq i\leq n$. Thus, it establishes the projections. In our setting,
we get the projections from the identity morphisms together with weakening. We
already have weakening, so here it suffices to introduce the identity morphisms.
\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}}
& & \inference{\jfameq{\Gamma}{A}{A'}}{\jtermeq{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}{\idfunc[A']}}
\end{align}
Identity functions are determined by their behavior with respect to substitution combined with
weakening. The identity functions will also be subject to compatibility rules.
\begin{align}
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{x}{\idfunc[A]}}{x}}\label{idfunc-subst-defn}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{A}}{\subst{\idfunc[A]}{\ctxwk{A}{P}}}{P}}\label{idfunc-wk-defn}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{A}}{\subst{\idfunc[A]}{\ctxwk{{A}{A}}{P}}}{P}}\label{idfunc-wk-defn2}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\idfunc[A]}{\ctxwk{A}{f}}}{f}}\label{idfunc-precomp} \\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\idfunc[A]}{\ctxwk{{A}{A}}{f}}}{f}}\label{idfunc-precomp} \\
& \inference{\jterm{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{\subst{g}{\ctxwk{B}{\idfunc[A]}}}{g}}\label{idfunc-postcomp}
\end{align}

@@

\begin{lem}
The following inference is valid:
\begin{equation*}
\inference{\jfam{\ctxext{\Gamma}{A}}{P}\quad\jfam{\ctxext{\Gamma}{A}}{Q}}
{\jfameq{\ctxext{\Gamma}{{A}{P}}}{\subst{\cprojfstf{A}{P}}{\ctxwk{\ctxext{A}{P}}{Q}}}{\ctxwk{P}{Q}}}
\end{equation*}
\end{lem}

\begin{proof}
Let $\jfam{\ctxext{\Gamma}{A}}{P}$ and $\jfam{\ctxext{\Gamma}{A}}{Q}$ be
families. Then we have
\begin{align*}
\subst{\cprojfstf{A}{P}}{\ctxwk{\ctxext{A}{P}}{Q}} 
& \jdeq \subst{\ctxwk{P}{\idfunc[A]}}{\ctxwk{\ctxext{A}{P}}{Q}} \tag{by definition}\\
& \jdeq \subst{\ctxwk{P}{\idfunc[A]}}{\ctxwk{P}{{A}{Q}}} \tag{by \autoref{comp-ww-f}}\\
& \jdeq \ctxwk{P}{\subst{\idfunc[A]}{\ctxwk{A}{Q}}} \tag{by \autoref{comp-ws-f}}\\
& \jdeq \ctxwk{P}{Q} \tag{by \autoref{idfunc-wk-defn}}
\end{align*}
\end{proof}

@@

\begin{proof}
To prove the first judgmental equality, note that
\begin{align*}
w & \jdeq \subst{w}{\idfunc[\ctxext{A}{P}]} \tag{by \autoref{idfunc-subst-defn}}\\
& \jdeq \subst{w}{\subst{\idfunc[P]}{{\ctxwk{P}{\idfunc[A]}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}
  \tag{by \autoref{idfunc-ext-comp}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{\subst{w}{{\ctxwk{P}{\idfunc[A]}}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}\tag{by \autoref{comp-ss-t}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\subst{w}{\idfunc[\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}]}}}\tag{by \autoref{comp-ss-t}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\subst{w}{\ctxwk{\ctxext{A}{P}}{\idfunc[\ctxext{A}{P}]}}}}\tag{by \autoref{idfunc-wk-comp}}\\
& \jdeq \subst{{w}{\idfunc[P]}}{{\subst{w}{\ctxwk{P}{\idfunc[A]}}}{\idfunc[\ctxext{A}{P}]}}\tag{by \autoref{defn-ws-4}}\\
& \jdeq \ctxext{\cprojfst{A}{P}{w}}{\cprojsnd{A}{P}{w}}\tag{by definition}
\end{align*}
\end{proof}

@@

\begin{lem}
When we substitute an extended term $\ctxext{f}{g}$ of $\ctxext{P}{Q}$ by a term
$x$ of $A$, the term that we get is $\ctxext{\subst{x}{f}}{\subst{x}{g}}$.
More precisely, the following inference rules are valid:
\begin{align*}
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}\quad\jterm{\ctxext{\Gamma}{A}}{\subst{f}{Q}}{g}}{\jtermeq{\Gamma}{\ctxext{\subst{x}{P}}{\subst{x}{Q}}}{\subst{x}{\ctxext{f}{g}}}{\ctxext{\subst{x}{f}}{\subst{x}{g}}}}\\
& \inference{\jterm{\ctxext{{\Gamma}{A}}{P}}{Q}{g}\quad\jterm{\ctxext{{\Gamma}{A}}{P}}{\subst{g}{R}}{t}}{\jtermeq{\ctxext{\Gamma}{\subst{x}{P}}}{\ctxext{\subst{x}{Q}}{\subst{x}{R}}}{\subst{x}{\ctxext{g}{t}}}{\ctxext{\subst{x}{g}}{\subst{x}{t}}}}
\end{align*}
\end{lem}
Finally, when we substitute by an extended term we get an equal result as when we
substitute two consecutive times. This looks like currying.

@@

\begin{proof}
We prove only the first judgmental equality. All the others are similar.
Let $\jterm{\Gamma}{A}{x}$ and $\jterm{\Gamma}{\subst{x}{P}}{u}$
be terms and let $\jfam{\ctxext{{\Gamma}{A}}{P}}{Q}$ be a family. Then we have
\begin{align*}
\subst{\ctxext{x}{u}}{Q} 
& \jdeq \subst{{u}{{x}{\idfunc[\ctxext{A}{P}]}}}{Q} \tag{by definition}\\
& \jdeq \subst{{u}{{x}{\idfunc[\ctxext{A}{P}]}}}{{x}{\ctxwk{A}{Q}}}\tag{by \autoref{defn-ws-3}}\\
& \jdeq \subst{{u}{{x}{\idfunc[\ctxext{A}{P}]}}}{{u}{\ctxwk{\subst{x}{P}}{\subst{x}{\ctxwk{A}{Q}}}}}\tag{by \autoref{defn-ws-3}}\\
& \jdeq \subst{{u}{{x}{\idfunc[\ctxext{A}{P}]}}}{{u}{{x}{\ctxwk{P}{{A}{Q}}}}}\tag{by \autoref{comp-sw-f}}\\
& \jdeq \subst{u}{{{x}{\idfunc[\ctxext{A}{P}]}}{{x}{\ctxwk{{P}{{A}{Q}}}}}}\tag{by \autoref{comp-ss-f}}\\
& \jdeq \subst{u}{{x}{{\idfunc[\ctxext{A}{P}]}{\ctxwk{P}{{A}{Q}}}}}\tag{by \autoref{comp-ss-f}}\\
& \jdeq \subst{u}{{x}{{\idfunc[\ctxext{A}{P}]}{\ctxwk{\ctxext{A}{P}}{Q}}}}\tag{by \autoref{comp-ew-f}}\\
& \jdeq \subst{u}{{x}{Q}}\tag{by \autoref{idfunc-wk-defn}}
\end{align*}
\end{proof}
